Toronto's Most Studious Wards:  An Analysis of Annual Toronto Public Library's Circulation and Visits by Wards Prior to and Post-COVID

### 2024-01-24 outputs/paper/paper.qmd
final paper upload
paper upload

Basic data cleaning was first applied to the column value names in order to ensure readability. As this is a report focused on investigating annual circulation and visits data by ward and year, it was decided that rows of branches that had a value of `0` in the ‘physical_branch’ which indicated that they did not have a physical location that patrons could enter were removed from the data set. The reason for this removal was to focus on public library branches that were physically accessible to patrons. Additionally, `branch_code`, `ward_name`, and `ward_no` were the only column values that were kept in the cleaned data set. This is because the purpose of this data set was to provide a way for the branch codes present in the *Library Circulation* and *Library Visits* data sets to be assigned their respective ward number for graphical analysis. The final cleaned version of the *Library Branch General Information* the data set has these changes reflected. (@tbl-cleaned_branch_data)

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-cleaned_branch_data
#| tbl-cap: Sample of Cleaned Library Branch Data

head(cleaned_branch_data, 8) |>
  kable(
    col.names = c("Branch Code", "Ward Number", "Ward Name"),
    booktabs = TRUE
  )
```

@@

Basic cleaning was first applied to the column value names to ensure readability as a CSV file. Similar to what was done for the cleaning of *Library Branch General Information*, non-physical branches were removed from the data set in addition to the branch code of `VR` which represented electronic circulation totals. This is because the scope of the report focuses only on physical branches that are accessible to patrons. Rows that did not contain `year` values between 2012 and 2017 were also removed since the scope of this report focuses on a five-year span between 2018 and 2022. The `subset()` function was used to remove rows containing multiple argument values. [@subsetfunction] The aforementioned changes were reflected in the cleaned data set. (@tbl-cleaned_circulation_data)

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-cleaned_circulation_data
#| tbl-cap: Sample of Cleaned Circulation Data

head(cleaned_circulation_data, 8) |>
  kable(
    col.names = c("Year", "Branch Code", "Total Circulation"),
    booktabs = TRUE
  )
```

@@

Basic cleaning was applied to this data set’s column value names to ensure readability. Following what was done on the two previous data sets, non-physical branches were removed from the data set since the scope of the report is focused on physical branches accessible to patrons. Additionally, the `BKONE` branch code was removed since it represented the annual visits of the two bookmobiles. The reason for this removal was due to its lack of fixed physical presence which could be found within one specific ward in Toronto. Rows that did not contain `year` values between 2012 and 2017 were also removed since the scope of this report focuses on five years between 2018 and 2022. The `subset()` function was used to remove rows containing multiple argument values. [@subsetfunction] These changes to the raw data set are viewable in the cleaned table. (@tbl-cleaned_visits_data)

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-cleaned_visits_data
#| tbl-cap: Sample of Cleaned Visits Data

head(cleaned_circulation_data, 8) |>
  kable(
    col.names = c("Branch Code", "Year", "Total Visits"),
    booktabs = TRUE
  )
```

### Annual Library Circulation and Visits by Ward
To get data sets that feature the annual circulation and visits total for each ward, the cleaned library branch data set was appended to each respective cleaned data set using the `merge()` function. Data sets were merged by `branch_code` which allowed for the corresponding ward numbers found in the library branch data set to be appended to the rows of annual total data that are sorted by branch code. [@mergedataframes] After the data sets were merged, they were grouped by ward number and then a year to be summarized by total visits or circulation. This process resulted in a data set producing the total annual circulation number for each ward and year between 2018 and 2022 as seen in its table. (@tbl-branch_circulation_data)  It also produced a data set producing the total annual visit number for each ward and year between 2018 and 2022 as seen in its resulting table. (@tbl-branch_visits_data) The reason for creating these data sets was to be able to produce graphs which visualize annual circulation and visit totals by ward instead of branches.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-branch_circulation_data
#| tbl-cap: Sample of Merged Annual Circulation Data by Ward

head(branch_circulation_data, 8) |>
  kable(
    col.names = c("Ward Number", "Year", "Total Circulation"),
    booktabs = TRUE
  )
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-branch_visits_data
#| tbl-cap: Sample of Merged Annual Vists Data by Ward

head(branch_circulation_data, 8) |>
  kable(
    col.names = c("Ward Number", "Year", "Total Visits"),
    booktabs = TRUE
  )
```

### City-Wide Annual Circulation and Visits
Based on the two previous data sets which aim to display total annual circulation and total by ward, they were also used to create city-wide totals for each year. The data sets were grouped by year and ward totals of circulation and visits were summed. This resulted in one data set which provides the total annual circulation of the Toronto Public Library’s physical and fixed location branches (@tbl-tpl_annual_circulation)  in addition to another data set which provides the total annual visits of the Toronto Public Library’s physical and fixed location branches. (@tbl-tpl_annual_visits)

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-tpl_annual_circulation
#| tbl-cap: Total Annual Circulation of Toronto Public Library Materials (Physical and Fixed Location Branches Only) Data Set

head(annual_circulation_data, 5) |>
  kable(
    col.names = c("Year", "Total Circulation"),
    booktabs = TRUE
  )
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-tpl_annual_visits
#| tbl-cap: Total Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) Data Set

head(annual_visits_data, 5) |>
  kable(
    col.names = c("Year", "Total Visits"),
    booktabs = TRUE
  )
```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-circulation_2018
#| fig-cap: Total 2018 Annual Circulation of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-circulation_2019
#| fig-cap: Total 2019 Annual Circulation of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-circulation_2020
#| fig-cap: Total 2020 Annual Circulation of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-circulation_2021
#| fig-cap: Total 2021 Annual Circulation of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-circulation_2022
#| fig-cap: Total 2022 Annual Circulation of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

As seen in a broad view of @fig_circulation_2018, @fig_circulation_2019, @fig_circulation_2020, @fig_circulation_2021, and @fig_circulation_2022 as a set, there appears to be a sharp decrease in library circulation starting in 2020. In 2018 and 2019, the mean annual circulation across wards was 910130.2 and 876039.6 respectively. This is comparatively higher than 2020 which has a mean of 	
409095.4. This mean was higher in 2021 at 517534.6 transactions. There was also an increase in 2022 at 620818.8.

Between 2018 and 2020, it appears that the total number of circulation transactions was trending towards an annual decrease which is reflected in @fig_circulation_2018, @fig_circulation_2019, and @fig_circulation_2020 when viewing the annual graphs as a set. Looking at the individual ward totals of circulation in these graphs reveals an overall decrease for each ward between these years. Similarly, between 2020 and 2022, there appears to be a positive trend in total circulation by ward when viewing the bar changes for each ward through @fig_circulation_2020, @fig_circulation_2021, and @fig_circulation_2022.

In 2018 and 2019, Ward 17 - Don Valley North, had the highest total circulation transactions. Libraries found in Don Valley North are Ethennonnhawahstihnen', Fairview, Hillcrest, and Pleasant View. For the same years, Ward 5 - York South-Weston had the lowest total circulation transactions. Libraries found in York South-Weston are Amesbury Park, Evelyn Gregory, Mount Dennis, and Weston. These ward totals are viewable in @fig_circulation_2018 and @fig_circulation_2019.

Between 2020 and 2023, Ward 4 - Parkdale-High Park had the highest total circulation transactions. Libraries found in Ward 4 include Annette Street, High Park, Jane/Dundas, Parkdale, Runnymede, and Swansea Memorial. Ward 7 - Humber River-Black Creek had the lowest total circulation transactions during this period. Libraries found in Ward 7 include Black Creek, Humber Summit, Jane/Sheppard, Woodville Park, and York Woods. These ward totals are viewable in @fig_circulation_2020, @fig_circulation_2021, and @fig_circulation_2022.

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-visits_2018
#| fig-cap: Total 2018 Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-visits_2019
#| fig-cap: Total 2019 Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-visits_2020
#| fig-cap: Total 2020 Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-visits_2021
#| fig-cap: Total 2021 Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-visits_2022
#| fig-cap: Total 2021 Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

As seen in a broad view of @fig_visits_2018, @fig_visits_2019, @fig_visits_2020, @fig_visits_2021, and @fig_visits_2022 as a set, there appears to be a sharp decrease in library circulation starting in 2020. In 2018 and 2019, the mean circulation across wards was 700570.1 and 687754.3 respectively. This is comparatively higher than 2020 which has a mean of 221601.5. This mean was even lower in 2021 at 160688.1. However, the mean circulation by ward increased in 2022 at 381105.3 after two consecutive years of decrease. Although 2022’s mean circulation by ward is an improvement to the two prior years, it is roughly only 50% of the means seen in 2018 and 2019.

Between 2018 and 2020, there appears to be a sharp decrease in total library visits between 2019 and 2020. This pattern is reflected in @fig_visits_2018, @fig_visits_2019, and @visits_2020 when viewing the bar graphs in a set. 2018 and 2019 have visually similar graphs with six wards having a visit total above 700,000. However, this is contrasted by @fig_visits_2020 where no ward has a visit total above 700,000. Interestingly, it appears in @fig_visits_2021 that visits to libraries across all wards were even lower than in 2020. In 2022, @fig_visits_2022 highlights total visits across all wards increasing which could indicate a future possible trend upwards after four consecutive years of decrease.

@@

### City-Wide Annual Circulation and Visits
Tables that provide a summary of the city-wide total visits (@tbl-tpl_annual_visits) and total circulation (tbl-tpl_annual_circulation) for each year. Annual circulation and visitation metrics have high standard deviations of 5502796 and 6349144, respectively. The high standard deviations imply that the overall annual variance of these metrics is high, therefore indicating values are spread out from the mean. This is seen in the graphs for both usage metrics as 2020 and 2021 reveal significantly smaller values compared to 2018 and 2019. By viewing the annual data, it is clear that the graphs represent the annual data after comparing the table to the aforementioned sections.

>>>>

Basic data cleaning was first applied to the column value names in order to ensure readability. As this is a report focused on investigating annual circulation and visits data by ward and year, it was decided that rows of branches that had a value of `0` in the ‘physical_branch’ which indicated that they did not have a physical location that patrons could enter were removed from the data set. The reason for this removal was to focus on public library branches that were physically accessible to patrons. Additionally, `branch_code`, `ward_name`, and `ward_no` were the only column values that were kept in the cleaned data set. This is because the purpose of this data set was to provide a way for the branch codes present in the *Library Circulation* and *Library Visits* data sets to be assigned their respective ward number for graphical analysis. The final cleaned version of the *Library Branch General Information* the data set has these changes reflected. (@tbl-cleanedbranchdata)

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-cleanedbranchdata
#| tbl-cap: Sample of Cleaned Library Branch Data

# Kable styling refernced from from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(cleaned_branch_data, 8) |>
  kable(
    col.names = c("Branch Code", "Ward Number", "Ward Name"),
    booktabs = TRUE
  )
```

@@

Basic cleaning was first applied to the column value names to ensure readability as a CSV file. Similar to what was done for the cleaning of *Library Branch General Information*, non-physical branches were removed from the data set in addition to the branch code of `VR` which represented electronic circulation totals. This is because the scope of the report focuses only on physical branches that are accessible to patrons. Rows that did not contain `year` values between 2012 and 2017 were also removed since the scope of this report focuses on a five-year span between 2018 and 2022. The `subset()` function was used to remove rows containing multiple argument values. [@subsetfunction] The aforementioned changes were reflected in the cleaned data set. (@tbl-cleanedcirculationdata)

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-cleanedcirculationdata
#| tbl-cap: Sample of Cleaned Circulation Data

# Kable styling refernced from from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(cleaned_circulation_data, 8) |>
  kable(
    col.names = c("Year", "Branch Code", "Total Circulation"),
    booktabs = TRUE
  )
```

@@

Basic cleaning was applied to this data set’s column value names to ensure readability. Following what was done on the two previous data sets, non-physical branches were removed from the data set since the scope of the report is focused on physical branches accessible to patrons. Additionally, the `BKONE` branch code was removed since it represented the annual visits of the two bookmobiles. The reason for this removal was due to its lack of fixed physical presence which could be found within one specific ward in Toronto. Rows that did not contain `year` values between 2012 and 2017 were also removed since the scope of this report focuses on five years between 2018 and 2022. The `subset()` function was used to remove rows containing multiple argument values. [@subsetfunction] These changes to the raw data set are viewable in the cleaned table. (@tbl-cleanedvisitsdata)

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-cleanedvisitsdata
#| tbl-cap: Sample of Cleaned Visits Data

# Kable styling refernced from from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(cleaned_circulation_data, 8) |>
  kable(
    col.names = c("Branch Code", "Year", "Total Visits"),
    booktabs = TRUE
  )
```

### Annual Library Circulation and Visits by Ward
To get data sets that feature the annual circulation and visits total for each ward, the cleaned library branch data set was appended to each respective cleaned data set using the `merge()` function. Data sets were merged by `branch_code` which allowed for the corresponding ward numbers found in the library branch data set to be appended to the rows of annual total data that are sorted by branch code. [@mergedataframes] After the data sets were merged, they were grouped by ward number and then a year to be summarized by total visits or circulation. This process resulted in a data set producing the total annual circulation number for each ward and year between 2018 and 2022 as seen in its table. (@tbl-branchcirculationdata) It also produced a data set producing the total annual visit number for each ward and year between 2018 and 2022 as seen in its resulting table. (@tbl-branchvisitsdata) The reason for creating these data sets was to be able to produce graphs which visualize annual circulation and visit totals by ward instead of branches.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-branchcirculationdata
#| tbl-cap: Sample of Merged Annual Circulation Data by Ward

# Kable styling refernced from from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(branch_circulation_data, 8) |>
  kable(
    col.names = c("Ward Number", "Year", "Total Circulation"),
    booktabs = TRUE
  )
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-branchvisitsdata
#| tbl-cap: Sample of Merged Annual Vists Data by Ward

# Kable styling refernced from from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(branch_circulation_data, 8) |>
  kable(
    col.names = c("Ward Number", "Year", "Total Visits"),
    booktabs = TRUE
  )
```

### City-Wide Annual Circulation and Visits
Based on the two previous data sets which aim to display total annual circulation and total by ward, they were also used to create city-wide totals for each year. The data sets were grouped by year and ward totals of circulation and visits were summed. This resulted in one data set which provides the total annual circulation of the Toronto Public Library’s physical and fixed location branches (@tbl-tplannualcirculation) in addition to another data set which provides the total annual visits of the Toronto Public Library’s physical and fixed location branches. (@tbl-tplannualvisits)

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-tplannualcirculation
#| tbl-cap: Total Annual Circulation of Toronto Public Library Materials (Physical and Fixed Location Branches Only) Data Set

# Kable styling refernced from from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(annual_circulation_data, 5) |>
  kable(
    col.names = c("Year", "Total Circulation"),
    booktabs = TRUE
  )
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-tplannualvisits
#| tbl-cap: Total Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) Data Set

# Kable styling refernced from from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(annual_visits_data, 5) |>
  kable(
    col.names = c("Year", "Total Visits"),
    booktabs = TRUE
  )
```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-circulation2018
#| fig-cap: Total 2018 Annual Circulation of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-circulation2019
#| fig-cap: Total 2019 Annual Circulation of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-circulation2020
#| fig-cap: Total 2020 Annual Circulation of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-circulation2021
#| fig-cap: Total 2021 Annual Circulation of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-circulation2022
#| fig-cap: Total 2022 Annual Circulation of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

As seen in a broad view of @fig-circulation2018, @fig-circulation2019, @fig-circulation2020, @fig-circulation2021, and @fig-circulation2022 as a set, there appears to be a sharp decrease in library circulation starting in 2020. In 2018 and 2019, the mean annual circulation across wards was 910130.2 and 876039.6 respectively. This is comparatively higher than 2020 which has a mean of 	
409095.4. This mean was higher in 2021 at 517534.6 transactions. There was also an increase in 2022 at 620818.8.

Between 2018 and 2020, it appears that the total number of circulation transactions was trending towards an annual decrease which is reflected in @fig-circulation2018, @fig-circulation2019, and @fig-circulation2020 when viewing the annual graphs as a set. Looking at the individual ward totals of circulation in these graphs reveals an overall decrease for each ward between these years. Similarly, between 2020 and 2022, there appears to be a positive trend in total circulation by ward when viewing the bar changes for each ward through @fig-circulation2020, @fig-circulation2021, and @fig-circulation2022.

In 2018 and 2019, Ward 17 - Don Valley North, had the highest total circulation transactions. Libraries found in Don Valley North are Ethennonnhawahstihnen', Fairview, Hillcrest, and Pleasant View. For the same years, Ward 5 - York South-Weston had the lowest total circulation transactions. Libraries found in York South-Weston are Amesbury Park, Evelyn Gregory, Mount Dennis, and Weston. These ward totals are viewable in @fig-circulation2018 and @fig-circulation2019.

Between 2020 and 2023, Ward 4 - Parkdale-High Park had the highest total circulation transactions. Libraries found in Ward 4 include Annette Street, High Park, Jane/Dundas, Parkdale, Runnymede, and Swansea Memorial. Ward 7 - Humber River-Black Creek had the lowest total circulation transactions during this period. Libraries found in Ward 7 include Black Creek, Humber Summit, Jane/Sheppard, Woodville Park, and York Woods. These ward totals are viewable in @fig-circulation2020, @fig-circulation2021, and @fig-circulation2022.

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-visits2018
#| fig-cap: Total 2018 Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-visits2019
#| fig-cap: Total 2019 Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-visits2020
#| fig-cap: Total 2020 Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-visits2021
#| fig-cap: Total 2021 Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-visits2022
#| fig-cap: Total 2022 Annual Visits of Toronto Public Library Branches (Physical and Fixed Location Branches Only) by Ward

@@

As seen in a broad view of @fig-visits2018, @fig-visits2019, @fig-visits2020, @fig-visits2021, and @fig-visits2022 as a set, there appears to be a sharp decrease in library circulation starting in 2020. In 2018 and 2019, the mean circulation across wards was 700570.1 and 687754.3 respectively. This is comparatively higher than 2020 which has a mean of 221601.5. This mean was even lower in 2021 at 160688.1. However, the mean circulation by ward increased in 2022 at 381105.3 after two consecutive years of decrease. Although 2022’s mean circulation by ward is an improvement to the two prior years, it is roughly only 50% of the means seen in 2018 and 2019.

Between 2018 and 2020, there appears to be a sharp decrease in total library visits between 2019 and 2020. This pattern is reflected in @fig-visits2018, @fig-visits2019, and @fig-visits2020 when viewing the bar graphs in a set. 2018 and 2019 have visually similar graphs with six wards having a visit total above 700,000. However, this is contrasted by @fig-visits2020 where no ward has a visit total above 700,000. Interestingly, it appears in @fig-visits2021 that visits to libraries across all wards were even lower than in 2020. In 2022, @fig-visits2022 highlights total visits across all wards increasing which could indicate a future possible trend upwards after four consecutive years of decrease.

@@

### City-Wide Annual Circulation and Visits
Tables that provide a summary of the city-wide total visits (@tbl-tplannualvisits) and total circulation (@tbl-tplannualcirculation) for each year. Annual circulation and visitation metrics have high standard deviations of 5502796 and 6349144, respectively. The high standard deviations imply that the overall annual variance of these metrics is high, therefore indicating values are spread out from the mean. This is seen in the graphs for both usage metrics as 2020 and 2021 reveal significantly smaller values compared to 2018 and 2019. By viewing the annual data, it is clear that the graphs represent the annual data after comparing the table to the aforementioned sections.
